Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study (draft only, not to be quoted)

### 2020-06-05 casecontrol.Rmd
last version before shortening pharmacoepi paper for NEJM
Fixed to drop controls no longer current

# Tables

```{r demog, echo=FALSE, warning=FALSE, message=FALSE}
options(knitr.kable.NA = '')

@@

knitr::kable(table.icdsubchapters,
             escape=FALSE, 
             booktabs=TRUE,
			 longtable=TRUE, 
             label="icdsubchapters",
             col.names=c(clean.header(colnames(table.icdsubchapters)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Univariate associations of severe disease with hospital diagnoses by ICD subchapters in those without any listed conditions: rows retained are those with p < 0.001 and at least 50 cases and controls") %>%
    column_spec(column=1, width="13cm") %>%
    column_spec(column=2:3, width="1.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    landscape() %>%
    kable_styling(latex_options="hold_position")

>>>>

# Tables
```{r demog, echo=FALSE, warning=FALSE, message=FALSE}
options(knitr.kable.NA = '')

@@

knitr::kable(table.icdsubchapters[grep("times", table.icdsubchapters$u.pvalue), ],
             escape=FALSE, 
             booktabs=TRUE,
			 longtable=TRUE, 
             label="icdsubchapters",
             col.names=c(clean.header(colnames(table.icdsubchapters)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Univariate associations of severe disease with hospital diagnoses by ICD subchapters in those without any listed conditions: rows retained are those with p < 0.001 and at least 50 cases and controls") %>%
    column_spec(column=1, width="13cm") %>%
    column_spec(column=2:3, width="1.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    landscape() %>%
    kable_styling(latex_options="hold_position")

### 2020-06-05 pharmaco.Rmd
Fixed to drop controls no longer current

---
title: 'Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study'
author:

@@

# Methods
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].  In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  The ascertainment of cases and controls is refreshed every few weeks, For this study, the accrued number of cases had increased to `r nrow(cc.severe[cc.severe$CASE==1, ])` For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.  Socioeconomic status was encoded as quintile of the postcode-based Scottish Index of Multiple Deprivation (SIMD).

@@

We have briefly reported associations of COVID-19 with dispensed prescriptions of drugs acting on other systems of interest.  The strongest independent effect of any class of drug acting on the cardiovascular system is that of loop diuretics.  This is likely to reflect confounding by heart failure as one of the main indications for use of loop diuretics. For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020]. A more detailed study of the relation of COVID-19 to cardiovascular disease and drugs prescribed for cardiovascular disease will be reported elsewhere.  The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].  Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was was markedly reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.  The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be accurately estimated.  As this drug is commonly used in combination with non-steroidal anti-inflammatory agents with which proton pump inhibitors are frequently co-prescribed, we would expect it to be associated with elevated risk in univariate analyses.

## Methodological strengths and weaknesses
A strength of this study is that it is based on dispensed prescriptions, rather than issued prescriptions.  Our study does not capture all exposure to drugs that are available over the counter, which include proton pump inhibitors.  Unless this misclassification of exposed individuals as unexposed differs between those not destined and those destined to develop severe COVID-19, it will lead to underestimation of the rate ratio associated with exposure to this class of drugs and underestimation of the attributable fraction.  Although we do not have morbidity data from primary care which would include risk factors such as smoking and obesity, we have diagnoses from hospital discharge records that have been coded to the ICD-10 classification by trained coders.  This diagnostic coding is more relevant to disease classification than the SNOMED diagnostic codes used in primary care databases. Although this case-control study is matched on primary care practice, in Scotland there is ample variation in high-risk prescribing between practitioners in the same practice [@guthrie_bad_2015], allowing study of the relationship of drug prescribing to COVID-19.

## Conclusion
Severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity, and it is strongest in those without hospital diagnoses of conditions that confer high risk of disease.  We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of all prescribed drugs. The profile of drug prescribing associated with COVID-19 includes many drug classes that are indicators of overprescribing and inappropriate polypharmacy.  The single drug class with the greatest explanatory power for the association of COVID-19 with polypharmacy is proton pump inhibitors.  Although this study did not identify unequivocal evidence of causality for the association with this drug class, effects on susceptibility to enteric infection with SARS-CoV-2 are biologically plausible for these drugs.  We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors, and should advise against their use without a recognized medical indication.

>>>>

---
title: 'Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study (draft only, not to be quoted)'
author:

@@

# Methods
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].  In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  The ascertainment of cases and controls is refreshed every few weeks, For this study, the accrued number of cases had increased to `r nrow(cc.severe[cc.severe$CASE==1, ])` For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive [and had not embarked -- to be checked in the CHI database by SW]. Socioeconomic status was encoded as quintile of the postcode-based Scottish Index of Multiple Deprivation (SIMD).

@@

We have briefly reported associations of COVID-19 with dispensed prescriptions of drugs acting on other systems of interest. We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of all prescribed drugs. For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020].  The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].  Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was was markedly reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.  The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be estimated reliably.

## Methodological strengths and weaknesses
A strength of this study is that it is based on dispensed prescriptions, rather than issued prescriptions.  Our study does not capture all exposure to drugs that are available over the counter, which include proton pump inhibitors.  Unless this misclassification of exposed individuals as unexposed differs between those not destined and those destined to develop severe COVID-19, it will lead to underestimation of the rate ratio associated with exposure to this class of drugs and underestimation of the attributable fraction.  Although we do not have morbidity data from primary care which would include risk factors such as smoking and obesity, we have diagnoses from hospital discharge records that have been coded to the ICD-10 classification by trained coders.

## Conclusion
Severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity, and it is strongest in those without hospital diagnoses of conditions that confer high risk of disease.  The profile of drug prescribing associated with COVID-19 includes many drug classes that are indicators of overprescribing and inappropriate polypharmacy.  The single drug class with the greatest explanatory power for the association of COVID-19 with polypharmacy is proton pump inhibitors.  Although this study did not identify unequivocal evidence of causality for this association, effects of this drug class on susceptibility to enteric infection with SARS-CoV-2 are biologically plausible.  We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors and on the avoidance of inappropriate polypharmacy.
